---
title: "AvianMetaNet: L0 File Preparation"
author: 
  - Phoebe L. Zarnetske
  - Patrick Bills
  - Emily Parker
  - Kelly Kapsar
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
editor: source
date: 10 February 2026
date-format: iso
---

## About

-   PROJECT: Ecological MetaNetworks
-   AUTHORS: Phoebe Zarnetske, Pat Bills, Emily Parker, Kelly Kapsar
-   COLLABORATORS: Vincent Miele, Stephane Dray
-   DATA INPUT: Data entry CSV files in L0 folder
-   DATA OUTPUT: L0 data: amn_all_raw.csv
-   DATE: 20 Mar 2023 - 19 FEB 2026
-   Next script to run: R/L1/3_subset_species_lists.R

**Run Date** 

```{r}

#| echo: false
#| warnings: false
print(Sys.Date())

```

### Introduction

This reads the various directories for CSVs in different stages, checks the files and combines them. Issues with files or data are reported, and total counts of files and rows added are listed. There are different folders checked here in different phases of

Before stitching files together, review and correct raw data files (e.g. species CSV file), using the notebook L0_file_review.rmd

The code here saves the file in the Data folder in L0 in the configuration of `filepaths.R`. see below

### Setup

*First set the file to save*

```{r}
stitched_L0_file <- "amn_all_raw.csv" 
```

*Then, setup the folders and files from configuration and check them*

```{r}

#| echo: false
#| warnings: false


source(here::here('R/lib/shared_functions.R'))

# This uses functions in R/lib/shared_functions.R which also imports R/lib/config.R
# which sets the locations of the data folders. 

file_paths <- get_file_paths() 
```

**File paths used for this run:**

```{r, echo=FALSE,}
print(file_paths)
```

## Processing CSVs

There are three main data folders for data files in different states. For each state, collect the list of files and o

### 1 Fully Checked CSV files for a species. `intxnsL0sp`

```{r}
csv_file_group_name='checked'
csv_file_path = file.path(file_paths$L0, "species")
# filter or add in google drive files here
checked_file_list <- list.files(path = csv_file_path , pattern = ".*\\.csv", 
  full.names = TRUE) %>% sort()

paste(csv_file_path, ":", csv_file_group_name, "files to process", length(checked_file_list))
```

Test reading a single file from the list of checked files

```{r}
i = sample(length(checked_file_list), 1)
exampleL0csv <- checked_file_list[i]
d = read_and_amend(exampleL0csv,add_entry_file_column= TRUE, fix_errors_on_read = TRUE)
print(names(d))
```

```{r}
tibble(d)
```

Stitching Files together:

```{r}
# create an intxns object/list with different aspects saved to allow for analysis
intxnsL0sp <- L0_stitch(
                csv_file_list = checked_file_list, 
                csv_file_group_name = csv_file_group_name,
                add_entry_file_column = TRUE, 
                fix_errors_on_read = TRUE
                )

nrow(intxnsL0sp$intxns)
```
Identify and report on missing fields: 

```{r}
df <- intxnsL0sp$intxns

no_url_or_scholar <- df[is.na(df$sourceA_URL) &
                        is.na(df$sourceB_URL) & 
                        is.na(df$sourceC_URL) & 
                        is.na(df$sourceD_URL) & 
                        is.na(df$DatabaseSearchURL),]

na_files <- unique(no_url_or_scholar$entry_file)
```

* Number of rows missing a source URL/Google Scholar URL: `r nrow(no_url_or_scholar)`
* Number of unique files with at least one row missing a URL: `r length(unique(no_url_or_scholar$entry_file))`



Transform 'long' format along sources from entered wide format.\
See documentation for this function in shared_functions.R

N Rows after transform:

```{r}
intxnsL0sp$intxns <- sources_wide_to_long(intxnsL0sp$intxns)
# intxnsL0sp is a list with several items to track counts of things and files
# see shared_functions.R for details

nrow(intxnsL0sp$intxns)

```
*Optionally examine the full stitched data frame:*

```{r}
#| echo: false
tibble(intxnsL0sp$intxns)
```

```{r}
#| echo: false
paste("Count of species after binding:", as.character(intxnsL0sp$count))
```

Comparison of Pre- and Post-Binding Values:

```{r}
#| echo: false
print_binding_report(intxnsL0sp)
```

### 2 not fully Checked Species in BBS `intxnsL0sptemp`

Optional. Uncomment to run the unchecked species

```         
# csv_file_group_name='species_temp'
# csv_file_path = file.path(file_paths$L0, "species_temp")
# # filter or add in google drive files here
# temp_file_list <- list.files(path = csv_file_path , pattern = ".*\\.csv", full.names = TRUE)
# 
# paste(csv_file_path, ":", csv_file_group_name, "files to process",  length(temp_file_list)) 
```

```         

# intxnsL0sptemp <- L0_stitch(
#                     csv_file_list = temp_file_list, 
#                     csv_file_group_name=csv_file_group_name
#                     )
# paste("Count of species after binding",intxnsL0sptemp$count)
```

```{r,echo=FALSE}
# print_binding_report(intxnsL0sptemp)
```

### 3 Data files (species) to be reviewed or in the process of being reviewed

Optional

(variable named intxnsL0spir )

```         
# csv_file_group_name='species_in_review'
# csv_file_path = file.path(file_paths$L0, "species_in_review")
# # filter or add in google drive files here
# in_review_file_list <- list.files(path = csv_file_path , pattern = ".*\\.csv", 
#   full.names = TRUE)

# paste(csv_file_path, ":", csv_file_group_name, "files to process", length(in_review_file_list))
```

```         

# intxnsL0spir <- L0_stitch(
#                   csv_file_list = in_review_file_list, 
#                   csv_file_group_name=csv_file_group_name
#                   ) 
```

```
# paste("Count of species after binding:", as.character(intxnsL0spir$count))
```

```
# print_binding_report(intxnsL0spir)
```

Before saving the species to file, un-comment and run the following to merge the species and species_in_review interaction data into checked species. These are lists with the data frames in the element `intxns`

```{r}
# echo both in-review and checked
# intxnsL0 <-rbind(intxnsL0sp$intxns, intxnsL0spir$intxns)
# or, do not echo in-review, only checked

intxnsL0 <- intxnsL0sp$intxns

```

**Count of species in saved L0:**

```{r}
#| echo: false
 nrow(intxnsL0)
```

> Note that some taxa1 in a given taxa1 csv could also be other species because of entering many pair-wise interactions in, for example, mixed flock entry. Any duplicates will be omitted later.


## Change species name columns to taxa names so that it reflects the broad range of taxonomic resolutions of the data
```{r}
intxnsL0 <- intxnsL0 %>% 
  rename(
    taxa1_scientific = species1_scientific, 
    taxa2_scientific = species2_scientific, 
    taxa1_common = species1_common,
    taxa2_common = species2_common,
    effect_tx1_on_tx2 = effect_sp1_on_sp2,
    effect_tx2_on_tx1 = effect_sp2_on_sp1
  )
```

## Save Results to Disk

export the data to become the current L0 interaction data, overwrite any existing file

```{r}
#| echo: false

# this file name is set at the top of this script, BUT 
# change the name here to save a test file or overwrite previous version
# stitched_L0_file <- "AvianInteractionData_L0_stitch.csv" 
L0_file <- write.csv(intxnsL0, 
                     file.path(file_paths$L0, stitched_L0_file),
                     row.names = F, 
                     fileEncoding = "UTF-8")

sessionInfo()
```
